---
title: "Distribution Functions and Order Statistics"
author: "Meredith Franklin"
date: "1/15/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(car)
```

## Setup {.tabset}

PM522b R code to illustrate cdf, pmf, and pdf. We show calculating the area under curve, quantiles, order statistics and sampling distributions related to normal: t, chisq, F.

### CDF
We plot the theoretical CDF for the gamma distribution with $\alpha=2$ and $\beta=3$ and add an empirical CDF.
```{r cdf}
    set.seed(235)
    df<-data.frame(x=rgamma(40,2,scale=3))
    ggplot(df, aes(x)) + stat_ecdf(geom = "step")+
      theme_bw()+
      stat_function(fun=pgamma,color="blue",args=list(shape=2,scale=3))+
      labs(title="ECDF and theoretical CDF")
   
    
```

### PMF
For discrete random variables we can plot the probability mass function, $P(X=x)$. For the Poisson distribution $P(X=x|\lambda)=\frac{e^{-\lambda}\lambda^x}{x!}$. In other words, for a given X=x, y is the probability that a random value takes that value.

```{r pmf}
x <- 0:10
y <- dpois(x,2)
df2<-data.frame(x=0:10,y=dpois(0:10,2))
ggplot(df2, aes(x,y)) + 
  geom_point()+
      theme_bw()
```

### PDF

```{r pdf}
mean=100
sd=15
lb=80
ub=120
# Create vector of x's: 100 observations around the mean with given sdev
x <- seq(-4,4,length=100)*sd + mean
# Create vector of y's: 100 observations from normal distribution with given mean and sdev
hx <- dnorm(x,mean,sd)

plot(x, hx, type="n", xlab="IQ Values", ylab="",
     main="Normal Distribution", axes=FALSE)

i <- x >= lb & x <= ub
lines(x, hx)
polygon(c(lb,x[i],ub), c(0,hx[i],0), col="red") 

area <- pnorm(ub, mean, sd) - pnorm(lb, mean, sd)
result <- paste("P(",lb,"< IQ <",ub,") =",
                signif(area, digits=3))
mtext(result,3)
axis(1, at=seq(40, 160, 20), pos=0)


# Triangle distribution
# f(x)=2x
x <- c(-0.5, 0, 1, 1, 1.5)
y <- c(0, 0, 2, 0, 0)
lb=0
ub=0.75


plot(x, y, lwd = 3,frame = FALSE, type = "l")
segments(0.75,0,x1=0.75,y1=1.5, col="black",lwd=2)
segments(0,0,0.75,0,col="black",lwd=2)
# create verticies (0,0),(0,0),(0.75,1.5),(0.75,0)
x.vertex<-c(0)
y.vertex<-c(0)
x.vertex<-c(x.vertex,0)
x.vertex<-c(x.vertex,0.75,0.75)
y.vertex<-c(y.vertex,0)
y.vertex<-c(y.vertex,1.5,0)
# plot polygon
polygon(x.vertex,y.vertex,col="red")

```

### Quantiles

We introduce order statistics with the uniform distribution and quantiles.
The $\alpha^{th}$ quantile of a distribution with distribution function $F$ is the point $x_\alpha$ so that $F(x_\alpha) = \alpha$

```{r quant}
x=seq(1,5,length=200)
y=rep(1/4,200)
plot(x,y,type="l",xlim=c(0,6),ylim=c(0,0.4),lwd=2,col="red",ylab="p",main="Unif(1,5)")
polygon(c(1,x,5),c(0,y,0),col="lightgray")
lines(x,y,type="l",lwd=2,col="red")

# Random sample of n=20 from uniform distribution

x<-runif(100,1,5)
quantile(x)
quantile(x,seq(0, 1, by=.1))
median(x)

samp.iqr<-function(x){
  q<-as.vector(quantile(x,probs=c(0.25,0.75)))
  return(q[2]-q[1])
}

samp.iqr(x)

# Other plots based on quantiles
# Boxplot
boxplot(x, main="Random Sample Uniform Distribution n=100")

# QQ plots
# compare X~U(1,5) to normal quantiles

qqnorm(x, main="Normal QQ Plot")
qqline(x)

# compare to uniform(0,1)
y<-runif(100,0,1)
qqplot(y,x)

# car package makes nicer plots
qqPlot(x,"unif", main="QQ Plot")

par(mfrow=c(1,2),oma=c(0,0,2,0))
qqplot(y,x)
qqPlot(x,"unif")
title("QQ plot Uniform",outer=TRUE)
```

### Order Statistics

```{r order}
f <- function(x, mu=0, sigma=1) dnorm(x, mean=mu, sd=sigma)
F <- function(x, mu=0, sigma=1) pnorm(x, mean=mu, sd=sigma, lower.tail=FALSE)

integrand <- function(x,j,n,mu=0, sigma=1) {
  x * (1 - F(x, mu, sigma))^(j-1) * F(x, mu, sigma)^(n-j) * f(x, mu, sigma)
}

E <- function(j,n, mu=0, sigma=1) {
  (1/beta(j,n-j+1)) * integrate(integrand,-Inf,Inf, j, n, mu, sigma)$value
}

E(1,1000)       # expected value of the minimum

E(1000,1000)    # expected value of the maximum

E(500.5,1000)   # expected value of the median

```


### T
The student's t-distribution is related to the normal distribution but when the variance is unknown. It has $p=n-1$ degrees of freedom and pdf $f_T(t)=\frac{\Gamma(\frac{p+1}{2})}{\Gamma(\frac{p}{2})(p\pi)^{1/2}}\frac{1}{(1+\frac{t^2}{p})^{(p+1)/2}}$

```{r t}

# Student's t distributions with various df and compare to the standard normal distribution

# Provide a range of values
x <- seq(-4, 4, length=100)

# different degrees of freedom
df <- c(1, 3, 5, 100)
# Set up colours and labels for the functions and plot legend
colors <- c("red", "blue", "darkgreen", "gold", "black")
labels <- c("df=1", "df=3", "df=5", "df=100", "standard normal")



# plot Z first
plot(x, dnorm(x), type="l", lty=4, lwd=4,xlab="x value",
     ylab="Density", main="t Distributions")

# add lines representing t-distribution
for (i in 1:4){
  lines(x, dt(x,df[i]), lwd=2, col=colors[i])
}

legend("topleft", labels, lwd=2, lty=c(1, 1, 1, 1, 2), col=colors,bty="n")

```

### Chi-square 
The chi-square distribution is related to the square of a standard normal distribution and has pdf $f(x)=\frac{1}{\Gamma(p/2)2^{p/2}}x^{(p/2)-1}e^{-x/2}$ for $p$ degrees of freedom.

```{r chisq}
x <- seq(0, 10, length=100)

# Plot chi-square with 1 df
plot(x, dgamma(x,shape=1,scale=2),type='l',lwd=6,lty=2,xlab="x value", ylab="Density", main="Chi-Square Distributions")

# Add lines for other df
df <- c(1, 2, 3,9)
# Set up colours and labels for the functions and plot legend
colors <- c("red","blue",  "orange", "darkgreen","black")
labels <- c("df=1", "df=2", "df=3", "df=9","gamma(1,2)")
for (i in 1:4){
  lines(x, dchisq(x,df[i]), lwd=2, col=colors[i])
}
legend("topright", labels, lwd=2, col=colors)
```

### F

The F distribution is used for comparing variances. It has pdf $f_F(f)=\frac{\Gamma(\frac{p+q}{2})}{\Gamma(\frac{p}{2})\Gamma(\frac{q}{2})}(\frac{p}{q})^{p/2}\frac{x^{(p/2)-1}}{[1+(p/q)x]^{(p+q)/2}}$ for $p$ and $q$ degrees of freedom.

```{r F}

x <- seq(0, 5, length=100)
# Plot F with 1 numerator and 5 denominator df
plot(x, df(x,1,5),type='l',lwd=2, main="F Distributions")
df1<-c(3,5,15)
df2<-c(5,10,25)
colors <- c("black","red", "blue", "orange")
# Add lines for additional numerator and denominator df
for (i in 1:3){
lines(x,df(x,df1[i],df2[i]),col=colors[i+1],lwd=2)
}

labels <- c("df=1,5", "df=3,10", "df=5,10", "df=15,25")
legend("topright", labels, lwd=2, col=colors)
```

